---
title: 'Assignment 7: GLMs week 2 (Linear Regression and beyond)'
author: "Monisha Eadala"
geometry: margin=2.54cm
output:
  word_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A06_GLMs_Week1.Rmd") prior to submission.

The completed exercise is due on Tuesday, February 25 at 1:00 pm.

## Set up your session 
1. Set up your session. Check your working directory, load the tidyverse, nlme, and piecewiseSEM packages, import the *raw* NTL-LTER raw data file for chemistry/physics, and import the processed litter dataset. You will not work with dates, so no need to format your date columns this time.

2. Build a ggplot theme and set it as your default theme.

```{r}
#1
getwd() # Checks my working directory

library(tidyverse)
library(nlme)
library(piecewiseSEM) # Loads the necessary packages

Lakes.ChemPhys <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")
Litter <- read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv") # Imports the necessary files

#2
theme_set(theme_classic()) # Builds a ggplot theme and sets it as my default theme
```


## NTL-LTER test
Research question: What is the best set of predictors for lake temperatures in July across the monitoring period at the North Temperate Lakes LTER? 

3. Wrangle your NTL-LTER dataset with a pipe function so that it contains only the following criteria: 

* Only dates in July (hint: use the daynum column). No need to consider leap years.
* Only the columns: lakename, year4, daynum, depth, temperature_C
* Only complete cases (i.e., remove NAs)

4. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature. Run a multiple regression on the recommended set of variables. 

```{r}
#3
Lakes.ChemPhys.July <- Lakes.ChemPhys %>% 
  filter(daynum > 181 & daynum <= 213) %>% # Keeps only dates in July
  select(lakename:daynum, depth, temperature_C) %>% # Keeps only the columns: lakename, year4, daynum, depth, temperature_C
  na.omit() # Removes NAs

#4
Temp.AIC <- lm(data = Lakes.ChemPhys.July, temperature_C ~ year4 + daynum + depth) # Runs an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature
step(Temp.AIC) # Confirms that I should be using all the variables since that model has the lowest AIC
Temp.model <- lm(data = Lakes.ChemPhys.July, temperature_C ~ year4 + daynum + depth) # Asserts the chosen model
summary(Temp.AIC) # Runs a multiple regression on the above recommended set of variables
```

5. What is the final set of explanatory variables that predict temperature from your multiple regression? How much of the observed variance does this model explain?

> Answer: The model with the lowest AIC is the final set of explanatory variables. Therefore, the final set of explanatory variables are year4, daynum and depth (as also mentioned above). The observed variance is the adjusted R-squared = 0.7398

6. Run an interaction effects ANCOVA to predict temperature based on depth and lakename from the same wrangled dataset.

```{r}
#6
Temp.ancova.interaction <- lm(data = Lakes.ChemPhys.July, temperature_C ~ lakename * depth)
summary(Temp.ancova.interaction) # Runs an interaction effects ANCOVA to predict temperature based on depth and lakename from the same wrangled dataset
```

7. Is there a significant interaction between depth and lakename? How much variance in the temperature observations does this explain?

> Answer: There is significant interaction between depth and lakeneme for all the lakes other than Paul Lake. Adjusted R-squared gives us the variance, which is 0.7855.

8. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r}
#8
Temp.plot <- ggplot(Lakes.ChemPhys.July, aes(x = depth, y = temperature_C, color = lakename)) + # Creates a graph that depicts temperature by depth, with a separate color for each lake
  geom_point(alpha = 0.5) + # Makes my points 50% transparent
  geom_smooth(method = "lm", se = FALSE) + # Adds a geom_smooth for each lake
  scale_color_brewer(palette = "Set1", direction = 1) + # Asserts a specific color palette that makes the graph effective and pretty
  ylim(0,35) + # Adjusts your y axis limits to go from 0 to 35 degrees
  labs(x = "Depth (m)", y = expression("Temperature ("*~degree*C*")"), color = "Lake Name") # Labels everything on the graph appropriately 
print(Temp.plot) 
```

9. Run a mixed effects model to predict dry mass of litter. We already know that nlcdClass and functionalGroup have a significant interaction, so we will specify those two variables as fixed effects with an interaction. We also know that litter mass varies across plot ID, but we are less interested in the actual effect of the plot itself but rather in accounting for the variance among plots. Plot ID will be our random effect.

a. Build and run a mixed effects model.
b. Check the difference between the marginal and conditional R2 of the model. 

```{r}
#9a
Litter.drymass.mixed <- lme(data = Litter, 
                            dryMass ~ nlcdClass*functionalGroup, 
                            random = ~1|plotID) # Builds a mixed effects model
summary(Litter.drymass.mixed) # Runs the mixed effects model

#9b
rsquared(Litter.drymass.mixed) # Gives me the marginal and conditional R2 of the model
# We see that the difference between the marginal and conditional R2 is 2.13%
```

b. continued... How much more variance is explained by adding the random effect to the model? 

> Answer: 2.13%  

c. Run the same model without the random effect.
d. Run an anova on the two tests. 
```{r}
#9c
Litter.drymass.fixed <- gls(data = Litter,
                      dryMass ~ nlcdClass*functionalGroup)
summary(Litter.drymass.fixed) # Runs the same model without the random effect

#9d
anova(Litter.drymass.mixed, Litter.drymass.fixed) # Runs an anova on the two tests
# The lower the AIC, the better
# The p-value tells us whether those models have a significantly different fit
```

d. continued... Is the mixed effects model a better model than the fixed effects model? How do you know?

> Answer: The mixed effects model is a better model than the fixed effects model since it has a lower AIC. The P-value is less than 0.0001; so it means that both the models have significantly different fit.