---
title: 'Assignment 3: Data Exploration'
author: "Monisha Eadala"
output:
  word_document: default
  pdf_document: default
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A03_DataExploration.Rmd") prior to submission.

The completed exercise is due on Tuesday, January 28 at 1:00 pm.


## Set up your R session

1. Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. 
```{r, message = FALSE} 
# To set up working directory
getwd()

# To load packages
library(tidyverse)

# To import datasets
Neonics <- read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv")
Litter <- read.csv("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv")

```

## Learn about your system

2. The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicologoy of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: We are interested in the ecotoxicologoy of neonicotinoids on insects since we want to know the effects of neonicotinoids on both target organisms and non-target organisms. From glancing at the Neonics dataset, we can tell that neonicotinoids seem to be effecting some of the beneficial insects and the harmful pests in a similar fashion (by causing mortality). For example, even though they seem to be causing mortality to the target pests such as coffee bean weevils and sweetpotato whiteflies, they are also causing the same harm to beneficial insects such as honey bees, bumble bees and few of the lady beetle species. Also it is necessary to have and understand such data that could aid in controlling the dossage and concentration of the pesticide according to the level of pest control required without going overboard causing harm to the ecological balance. 

3. The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: The litter and woody debris data may be used to estimate the annual Aboveground Net Primary Productivity (ANPP) and aboveground biomass at plot, site, and continental scales that are important to understand vegetative carbon fluxes over the time and answer very important questions regarding the global carbon balance, the location of the missing carbon sink, and predictions of global climate change. Such knowledge and predictions can influence current management practices and public policy.

4. How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here: 

> Answer: Litter and fine woody debris are collected individually from two different traps - elevated and ground traps. All masses reported after processing are reported at the spatial resolution of a single trap and the temporal resolution of a single collection event.
* Spatial Sampling: For elevated traps, litter and fine woody debris sampling is executed at terrestrial NEON sites that contain woody vegetation more than 2m tall. Locations of tower plots are selected randomly within the 90% flux footprint of the primary and secondary airsheds.
* In sites with forested tower airsheds, the litter sampling is targeted to take place in 20 40m x 40m plots. In sites with low-statured vegetation over the tower airsheds,litter sampling is targeted to take place in 4 40m x 40m tower plots (to accommodate co-located soil sampling) plus 26 20m x 20m plots. Also, Trap placement within plots may be either targeted or randomized, depending on the vegetation. In sites with more than 50% aerial cover of woody vegetation more than 2m in height, placement of litter traps is random.
* Temporal Sampling: Ground traps are sampled once per year. Target sampling frequency for elevated traps varies by vegetation present at the site, with frequent sampling (1x every 2weeks) in deciduous forest sites during senescence, and infrequent year-round sampling (1x every 1-2 months) at evergreen sites.


## Obtain basic summaries of your data (Neonics)

5. What are the dimensions of the dataset?

```{r}
dim(Neonics) # Gives the dimensions (number of rows and columns) of the dataset (in this the number of rows are 4623, while the number of columns are 30)

```

6. Using the `summary` function, determine the most common effects that are studied. Why might these effects specifically be of interest?
```{r}
summary(Neonics$Effect) # Lists all the effects with their respective frequency; the ones with the most frequency are the most common. 
```

> Answer: The most common effects that are studied are "population", "mortality", "behavior", "feeding behavior", "reproduction", and "development". These effects might specifically be of interest to understand what are some of the most common effects of the insecticides and to even determine if such effects are detrimental to the environment. Such understanding can form a basis for environmental decision analysis and can determine the future use of such insecticides.

7. Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r}
summary(Neonics$Species.Common.Name) # Lists the studied species with the most common ones listed at the top
```

> Answer: The six most commonly studies species in the dataset are "honey bee", "parasitic wasp", "buff tailed bumblebee", "Carniolan honey bee", and "bumble bee" and "Italian honeybee". the one thing all these species have in common is that they are all considered beneficial insects that feed on other pests that harm the crops. They all have good reputation in the field of garden and crop management/ pest control.

8. Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r}
class(Neonics$Conc.1..Author.) # Generates the class of Conc.1..Author.
```

> Answer: The class of Conc.1..Author. is "factor" and not numeric, and this is because the values cannot be considered numeric yet given that they all have different units. Only after converting all the units to a single consistent unit, we should convert the class of the variable to numeric.


## Explore your data graphically (Neonics)

9. Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year. 
```{r}
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year), bins = 50) # Generates a plot of the number of studies conducted by publication year
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.
```{r}
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year, color = Test.Location), bins = 50) # Reproduces the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common test locations overall according to the graph seem to be Lab, followed by Field natural, Field artifical and then Field underdeterminable. However, they do differ over the time. For instance, right before 2010 (around 2009) Field natural seems to be more common, while right after 2010 Lab seems to be more common than Field natural.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

```{r}
ggplot(Neonics, aes(x = Endpoint)) +
  geom_bar() # Creates a bar graph of Endpoint counts
```

> Answer: The two most common endpoints are NOEL (most common) and LOEL (second-most common). NOEL is the abbreviation for No-observable-effect-level, and it is defined as the highest dose (concentration) producing effects not significantly different from responses of controls according to author's reported statistical test (NOEAL/NOEC). LOEL is the abbreviation for Lowest-observable-effect-level, and it is definied as the lowest dose (concentration) producing effects that were significantly different (as reported by authors) from responses of controls (LOEAL/LOEC). 

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018. 
```{r}
class(Litter$collectDate) # Determines the class of collectDate (in this case it is a "factor")
Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y-%m-%d") # Changes it to a date
class(Litter$collectDate) # Confirmed that the new class of collectDate is a "Date"

```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
unique(Litter$plotID) # Determines how many plots were sampled at Niwot Ridge without duplication
```

> Answer: The 'unique' function eliminates duplicate elements/rows from a vector, data frame or array, while the 'summary' function does not. In other words, the 'unique' function strictly returns the unique rows in a dataframe while the summary returns all the rows in a dataframe. For the 'unique' function, all the elements in each row must match in order to be termed as duplicates.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.
```{r}
ggplot(Litter, aes(x = functionalGroup)) +
  geom_bar() # Creates a bar graph of functionalGroup counts
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup. 
```{r}
ggplot(Litter) +
  geom_boxplot(aes(x = functionalGroup, y = dryMass)) # Creates a boxplot of dryMass by functionalGroup

ggplot(Litter) +
  geom_violin(aes(x = functionalGroup, y = dryMass), 
              draw_quantiles = c(0.25, 0.5, 0.75)) # Creates a violin plot of dryMass by functionalGroup
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: A boxplot is a more effective visualization option than the violin plot in this case because the boxplot is giving us more information; especially regarding the median, first quartile, third quartile and outliers. On the other hand, the only information the violin plot is giving us is the minimum and the maximum observations (which the box plot gives us as well). Also, R tells us that there are 3 counts of collapsing due to unique 'x' values in the violin plot, which makes the visual even more unappealing. Therefore, a boxplot is much better in this case.

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: From the box plot it is evident that Needles tend to have the highest biomass at these sites. After Needles, Mixed litter seems to have the next highest biomass.